---
title: "Denver traffic accident exploratory analysis"
author: patrick charles  
output:  
    html_document:
        keep_md: true
---

```{r prereqs, message=FALSE, warning=FALSE, echo=FALSE}
  if(!require(dplyr)) install.packages("dplyr", dep=T)
  if(!require(ggplot2)) install.packages("ggplot2", dep=T)
  if(!require(curl)) install.packages("curl", dep=T)
  if(!require(leaflet)) install.packages("leaflet", dep=T)
  if(!require(ggmap)) install.packages("ggmap", dep=T)
  if(!require(gridExtra)) install.packages("gridExtra", dep=T)
```

# Traffic "Accidents" Involving Bicycles in Denver

## data acquisition

### data source - Denver Open Data

```{r}
  traffic_url <- "http://data.denvergov.org/download/gis/traffic_accidents/csv/traffic_accidents.csv"
```

### save a copy of the data source file offline
```{r eval=FALSE}
  download.file(url=traffic_url, destfile='data/denver_traffic_accidents.csv', method='wget')
```

### and, read from disk w/ transformation
```{r}
# direct read
#  traffic <- read.csv( curl(traffic_url), sep=",", header=T )

  traffic <- read.csv("data/denver_traffic_accidents.csv",
    sep=",", header=TRUE, stringsAsFactors=FALSE, 
    colClasses=c("OFFENSE_TYPE_ID"="factor", "OFFENSE_CATEGORY_ID"="factor",
                 "NEIGHBORHOOD_ID"="factor",
                 "DISTRICT_ID"="factor", "PRECINCT_ID"="factor",
#                 "FIRST_OCCURRENCE_DATE"="POSIXct", "REPORTED_DATE"="POSIXct", 
# "LAST_OCCURRENCE_DATE"="POSIXct", - usually blank in this dataset
                 "OFFENSE_CODE"="factor"
  ))
```

### remove points with wildly inconsistent longitudes
```{r}
  # a valid denver longitude is ~-104
  traffic <- traffic[!(traffic$GEO_LON > -100), ]
```

### create new columns with time components
```{r}
  traffic$FIRST_OCCURRENCE_DATE <-
    as.POSIXct(traffic$FIRST_OCCURRENCE_DATE,format="%Y-%m-%d %H:%M:%S")
  traffic$REPORTED_DATE <-
    as.POSIXct(traffic$REPORTED_DATE,format="%Y-%m-%d %H:%M:%S")

  traffic$OCCURRENCE_HOUR <-
    as.numeric(format(traffic$FIRST_OCCURRENCE_DATE, "%H")) +
    as.numeric(format(traffic$FIRST_OCCURRENCE_DATE, "%M")) / 60

  traffic$OCCURRENCE_DAY <- 
    as.numeric(format(traffic$FIRST_OCCURRENCE_DATE, "%u"))

  traffic$OCCURRENCE_MONTH <- 
    as.numeric(format(traffic$FIRST_OCCURRENCE_DATE, "%m"))

  traffic$OCCURRENCE_YEAR <- 
    as.numeric(format(traffic$FIRST_OCCURRENCE_DATE, "%Y"))
```

### convenient data subsets
```{r}
pedestrians <- traffic[traffic$PEDESTRIAN_IND %in% c(1), ]
bicycles <- traffic[traffic$BICYCLE_IND %in% c(1), ]
```

## exploratory analysis

### General shape of the dataframe
```{r}
  head(traffic, 1)
  str(traffic)
```

### Neighborhoods with the most accidents
```{r label='neighborhood_counts'}
  traffic %>% count(NEIGHBORHOOD_ID) %>% arrange(desc(n))

  ggplot(traffic) + geom_bar(aes(x=NEIGHBORHOOD_ID, fill=DISTRICT_ID)) +
    scale_x_discrete(limits = (traffic %>% count(NEIGHBORHOOD_ID)
                                       %>% arrange(n))$NEIGHBORHOOD_ID) +
    coord_flip() +
    labs(title="Accidents by Neighborhood and District",
      x="Neighborhood", y="Count", fill="District")
```

### Districts and precincts with the most accidents
```{r label='district_counts'}
  traffic %>% count(DISTRICT_ID) %>% arrange(desc(n))
  traffic %>% count(DISTRICT_ID, PRECINCT_ID) %>% arrange(desc(n))

  ggplot(traffic) + geom_bar(aes(x=DISTRICT_ID, fill=PRECINCT_ID)) +
    scale_x_discrete(limits = (traffic %>% count(DISTRICT_ID)
                                       %>% arrange(desc(n)))$DISTRICT_ID) +
    labs(title="Accidents by District and Precinct",
      x="District", y="Count", fill="Precinct")
```

### Accidents by Offense Type
```{r label='offense_type'}
  ggplot(traffic) + geom_bar(aes(x=OFFENSE_TYPE_ID)) +
    scale_x_discrete(limits = (traffic %>% count(OFFENSE_TYPE_ID)
                                       %>% arrange(n))$OFFENSE_TYPE_ID) +
    labs(title="Number of Accidents by Offense Type",
      x="Offense Type", y="Count") +
    coord_flip()
```

## Geo and Time Visualizations

```{r message=FALSE}
  library(ggmap)
  library(gridExtra)

denver_central <- get_map(location="Denver", zoom=13, source="osm")
denver_full <- get_map(location =
    c(mean(traffic$GEO_LON), mean(traffic$GEO_LAT)), 
  zoom=12, source="osm")
```

### All Traffic Accidents (Yellow)
```{r label='all_accidents', warning=FALSE}
  plot <-
    geom_point(aes(x=traffic$GEO_LON, y=traffic$GEO_LAT),
      data=traffic, col="#ffff88", alpha=0.2, size=1)

#  p1 <- ggmap(denver_full) + plot
#  p2 <- ggmap(denver_central) + plot
#  grid.arrange(p1, p2, nrow=1)

  ggmap(denver_full) + plot
  ggmap(denver_central) + plot
```

### All Accidents by Time
```{r label='all_by_time', warning=FALSE}
  g1 <- ggplot(data=traffic, aes(traffic$OCCURRENCE_HOUR)) +
         geom_histogram(breaks=seq(0, 24, by=1), col="yellow",
         fill="#ffff88", alpha=0.4) + 
         labs(x="Hour of Day", y="Number of Accidents")

  g2 <- ggplot(data=traffic, aes(traffic$OCCURRENCE_DAY)) +
         geom_histogram(breaks=seq(0, 7, by=1), col="yellow",
         fill="#ffff88", alpha=0.4) + 
         labs(x="Day of Week (Mon-Sun)")

  g3 <- ggplot(data=traffic, aes(traffic$OCCURRENCE_MONTH)) +
         geom_histogram(breaks=seq(0, 12, by=1), col="yellow",
         fill="#ffff88", alpha=0.4) + 
         labs(x="Month (Jan-Dec)")

  g4 <- ggplot(data=traffic, aes(traffic$OCCURRENCE_YEAR)) +
         geom_histogram(breaks=seq(2011, 2016, by=1), col="yellow",
         fill="#ffff88", alpha=0.4) + 
         labs(x="Year")

  grid.arrange(g1, g2, g3, g4, nrow=2)
```

### All Incidents Involving Bicycles (Red)

```{r label='bicycle_accidents', warning=FALSE}
  g1 <-
    geom_point(aes(x=traffic$GEO_LON, y=traffic$GEO_LAT),
      data=traffic, col="#ffff88", alpha=0.2, size=1)
  g2 <-
    geom_point(aes(x=traffic$GEO_LON, y=traffic$GEO_LAT),
      data=traffic, col=ifelse(traffic$BICYCLE_IND==1, "#ff0000", "#ffff88"),
      alpha=0.4, size=1.5)

  ggmap(denver_full) + g1 + g2
  ggmap(denver_central) + g1 + g2
```

### All Incidents Involving Bicycles by Time

```{r label='bicycle_by_time'}
  g1 <- ggplot(data=bicycles, aes(bicycles$OCCURRENCE_HOUR)) +
    geom_histogram(breaks=seq(0, 24, by=1), col="red",
                   fill="#ff5555", alpha=0.4) + 
    labs(x="Hour of Day", y="Number of Accidents Involving Cyclists (Red)")

  g2 <- ggplot(data=bicycles, aes(bicycles$OCCURRENCE_DAY)) +
    geom_histogram(breaks=seq(0, 7, by=1), col="red",
                   fill="#ff5555", alpha=0.4) + 
    labs(x="Day of Week (Mon-Sun)")

  g3 <- ggplot(data=bicycles, aes(bicycles$OCCURRENCE_MONTH)) +
    geom_histogram(breaks=seq(0, 12, by=1), col="red",
                   fill="#ff5555", alpha=0.4) + 
    labs(x="Month (Jan-Dec)")

  g4 <- ggplot(data=bicycles, aes(bicycles$OCCURRENCE_YEAR)) +
    geom_histogram(breaks=seq(2011, 2016, by=1), col="red",
                   fill="#ff5555", alpha=0.4) + 
    labs(x="Year")

  grid.arrange(g1, g2, g3, g4, nrow=2)
```

### Hit and Run Incidents Involving Bicycles (Black)

```{r label='bicycle_hit_run', warning=FALSE}
  g1 <- geom_point(aes(x=traffic$GEO_LON, y=traffic$GEO_LAT),
    data=traffic, col="#ffff88", alpha=0.2, size=1)

  g2 <- geom_point(data=bicycles, aes(x=bicycles$GEO_LON, y=bicycles$GEO_LAT),
    col=ifelse( (bicycles$OFFENSE_TYPE_ID=="traffic-accident-hit-and-run"),
    "#000000", "#ff0000"), alpha=0.4, size=1.5)

  ggmap(denver_full) + g1 + g2
  ggmap(denver_central) + g1 + g2
```

### All Hit and Run Incidents Involving Bicycles by Time

```{r label='bicycle_hit_run_time'}
  bicyclehitrun <-
    bicycles[bicycles$OFFENSE_TYPE_ID %in% c("traffic-accident-hit-and-run"), ]

  g1 <- ggplot() +
    geom_histogram(data=bicycles, aes(bicycles$OCCURRENCE_HOUR),
      breaks=seq(0, 24, by=1), col="red", fill="#ff5555", alpha=0.4) + 
    geom_histogram(data=bicyclehitrun, aes(bicyclehitrun$OCCURRENCE_HOUR),
      breaks=seq(0, 24, by=1), col="black", fill="#222222", alpha=0.4) + 
    labs(x="Hour of Day", y="Hit and Run Accidents on Cyclists (Black)")

  g2 <- ggplot() +
    geom_histogram(data=bicycles, aes(bicycles$OCCURRENCE_DAY),
      breaks=seq(0, 7, by=1), col="red", fill="#ff5555", alpha=0.4) + 
    geom_histogram(data=bicyclehitrun, aes(bicyclehitrun$OCCURRENCE_DAY),
      breaks=seq(0, 7, by=1), col="black", fill="#222222", alpha=0.4) + 
    labs(x="Day of Week (Mon-Sun)")

  g3 <- ggplot() +
    geom_histogram(data=bicycles, aes(bicycles$OCCURRENCE_MONTH),
      breaks=seq(0, 13, by=1), col="red", fill="#ff5555", alpha=0.4) + 
    geom_histogram(data=bicyclehitrun, aes(bicyclehitrun$OCCURRENCE_MONTH),
      breaks=seq(0, 13, by=1), col="black", fill="#222222", alpha=0.4) + 
    labs(x="Month (Jan-Dec)")

  g4 <- ggplot() +
    geom_histogram(data=bicycles, aes(bicycles$OCCURRENCE_YEAR),
      breaks=seq(2011, 2016, by=1), col="red", fill="#ff5555", alpha=0.4) + 
    geom_histogram(data=bicyclehitrun, aes(bicyclehitrun$OCCURRENCE_YEAR),
      breaks=seq(2011, 2016, by=1), col="black", fill="#222222", alpha=0.4) + 
    labs(x="Year")

  grid.arrange(g1, g2, g3, g4, nrow=2)

```

### Incidents Involving Cyclists - Hit and Run Probability

```{r}
  traffic_hours <- count(traffic, floor(OCCURRENCE_HOUR))
  colnames(traffic_hours) <- c("hour", "count")

  bicycles_hours <- count(bicycles, floor(OCCURRENCE_HOUR))
  colnames(bicycles_hours) <- c("hour", "count")

  bicyclehitrun_hours <- count(bicyclehitrun, floor(OCCURRENCE_HOUR))
  colnames(bicyclehitrun_hours) <- c("hour", "count")

  comparison <- bicyclehitrun_hours
  comparison$accidents <- bicycles_hours$count
  comparison$hitrun <- bicyclehitrun_hours$count
  comparison$probability <- bicyclehitrun_hours$count / bicycles_hours$count
  comparison

  ggplot(comparison, aes(hour, probability)) + geom_point(size=1) +
    geom_smooth(span=1.0) + 
  labs(title="Incidents Involving Cyclists - Hit and Run Probability",
    x="Hour of Day", y="Prevalence")
```

### General Hit and Run Probability
```{r warning=FALSE}
  traffichitrun <-
    traffic[traffic$OFFENSE_TYPE_ID %in% c("traffic-accident-hit-and-run"), ]

  traffic_hours <- count(traffic, floor(OCCURRENCE_HOUR))
  colnames(traffic_hours) <- c("hour", "count")

  traffichitrun_hours <- count(traffichitrun, floor(OCCURRENCE_HOUR))
  colnames(traffichitrun_hours) <- c("hour", "count")

  traffic_comparison <- traffichitrun_hours
  traffic_comparison$accidents <- traffic_hours$count
  traffic_comparison$hitrun <- traffichitrun_hours$count
  traffic_comparison$probability <-
    traffichitrun_hours$count / traffic_hours$count

  ggplot(traffic_comparison, aes(hour, probability)) + geom_point(size=1) +
    geom_smooth(span=1.0) + 
  labs(title="All Traffic Incidents - Hit and Run Probability",
    x="Time of Day", y="Prevalence")
```

### Accidents Involving Pedestrian (Purple)

```{r label='pedestrian_accidents', warning=FALSE}
    g1 <- geom_point(aes(x=traffic$GEO_LON, y=traffic$GEO_LAT), data=traffic,
      col="#ffff88", alpha=0.2, size=1)
    g2 <- geom_point(aes(x=traffic$GEO_LON, y=traffic$GEO_LAT), data=traffic,
      col=ifelse(traffic$PEDESTRIAN_IND==1, "#ff00ff", "#ffff88"), alpha=0.4,
      size=1.5)
  ggmap(denver_full) + g1 + g2
  ggmap(denver_central) + g1 + g2
```

### Hit and Run on Pedestrians (Black)
```{r label='pedestrian_hit_run', warning=FALSE}
  pedestrians <- traffic[traffic$PEDESTRIAN_IND %in% c(1), ]
  g1 <- geom_point(aes(x=traffic$GEO_LON, y=traffic$GEO_LAT), data=traffic,
    col="#ffff88", alpha=0.2, size=1) 
  g2 <- geom_point(data=pedestrians, aes(x=pedestrians$GEO_LON,
    y=pedestrians$GEO_LAT),
    col=ifelse( (pedestrians$OFFENSE_TYPE_ID=="traffic-accident-hit-and-run"),
    "#220022", "#ff00ff"), alpha=0.4, size=1.5, position="jitter")

  ggmap(denver_full) + g1 + g2
  ggmap(denver_central) + g1 + g2
```

## Summary Stats

### Earliest Incident
```{r}
  head(sort(traffic$FIRST_OCCURRENCE_DATE), 1)
```

### Latest Incident
```{r}
  tail(sort(traffic$FIRST_OCCURRENCE_DATE), 1)
```

### Total Number of Incidents
```{r}
  total_incidents <- nrow(traffic)
  total_incidents
```

### Number of Hit and Run Incidents
```{r}
  hit_and_run_incidents <-
    nrow(bicycles[traffic$OFFENSE_TYPE_ID %in%
      c("traffic-accident-hit-and-run"), ])
  hit_and_run_incidents
```

### Number of Vehicular Homicides
```{r}
  nrow(traffic[traffic$OFFENSE_TYPE_ID %in% c("traf-vehicular-homicide"), ])
```

### Number of Incidents on Bicycles
```{r}
  bicycle_incidents <- nrow(bicycles)
  bicycle_incidents
```

### Number of Incidents on Pedestrians
```{r}
  pedestrian_incidents <- nrow(pedestrians)
  pedestrian_incidents
```

### Number of Hit and Run Incidents on Bicycles
```{r}
  bicycle_hit_and_run_incidents <-
    nrow(bicycles[bicycles$OFFENSE_TYPE_ID %in%
      c("traffic-accident-hit-and-run"), ])
  bicycle_hit_and_run_incidents
```

### Number of Vehicular Homicides on Bicycles
```{r}
# aka "killing cyclists is always an 'accident'"
  nrow(bicycles[bicycles$OFFENSE_TYPE_ID %in% c("traf-vehicular-homicide"), ])
```

### Number of Hit and Run Incidents on Pedestrians
```{r}
  pedestrian_hit_and_run_incidents <-
    nrow(pedestrians[pedestrians$OFFENSE_TYPE_ID %in%
      c("traffic-accident-hit-and-run"), ])
  pedestrian_hit_and_run_incidents
```

### Cyclist Incidents, Percentage Hit and Run
```{r}
  cyclist_hit_and_run_percentage <-
    bicycle_hit_and_run_incidents * 100 / bicycle_incidents
  cyclist_hit_and_run_percentage
```

### Total Incidents, Percentage Hit and Run
```{r}
  hit_and_run_percentage <- hit_and_run_incidents * 100 / total_incidents
  hit_and_run_percentage
```

### Pedestrian Incidents, Percentage Hit and Run
```{r}
  pedestrian_hit_and_run_percentage <-
    pedestrian_hit_and_run_incidents * 100 / pedestrian_incidents
  pedestrian_hit_and_run_percentage
```

### Incidents on a Specific Street
```{r}
  length(which(grepl("COLFAX", bicycles$INCIDENT_ADDRESS)))
```

## Density Heatmaps

### Density - All Incidents
```{r label='density_all', warning=FALSE}
  ggmap(denver_full) +
    geom_density2d(data = traffic, aes(x=traffic$GEO_LON, y=traffic$GEO_LAT)) +
    stat_density2d(data = traffic, aes(x=traffic$GEO_LON, y=traffic$GEO_LAT,
      fill = ..level.., alpha = ..level..),
      size = 1.0, bins = 16, geom = 'polygon') +
   scale_fill_gradient(low = "green", high = "red") +
   scale_alpha(range = c(0.00, 0.4), guide = FALSE) 
```

### Density - Bicycle Incidents
```{r label='density_bicycle', warning=FALSE}
  ggmap(denver_full) +
    geom_density2d(data = bicycles, aes(x=bicycles$GEO_LON,
      y=bicycles$GEO_LAT)) +
   stat_density2d(data = bicycles, aes(x=bicycles$GEO_LON, y=bicycles$GEO_LAT,
     fill = ..level.., alpha = ..level..),
     size = 1.0, bins = 16, geom = 'polygon') +
   scale_fill_gradient(low = "green", high = "red") +
   scale_alpha(range = c(0.00, 0.4), guide = FALSE) 
```

### Interactive Map - Bicycle Incidents
```{r label='leaflet', eval=FALSE}

#  traffic_map <- leaflet() %>%
#    addTiles() %>%
#    addMarkers(lng=bicycles$GEO_LON, lat=bicycles$GEO_LAT)
#  traffic_map

  bicycle_interactive <- leaflet() %>%
    setView(lng=mean(bicycles$GEO_LON),
            lat=mean(bicycles$GEO_LAT), zoom=12) %>%
    addTiles() %>%
    addMarkers(lng=bicycles$GEO_LON, lat=bicycles$GEO_LAT,
      popup=paste(bicycles$FIRST_OCCURRENCE_DATE, bicycles$OFFENSE_TYPE_ID, "bicycle", bicycles$INCIDENT_ADDRESS, sep=", "))
  bicycle_interactive
```
